Bayesian inference of material culture phylogenies using continuous traits: A birth–death model for Late Neolithic/Early Bronze Age arrowheads from Northwestern Europe

### 2023-10-17 analysis/paper/paper.qmd
submit to SocArXiv

Bayesian inference consists of three parts: (1) likelihood models describing the processes that generated the data (statements of what the researcher believes to be true about their data), with the models containing (2) parameters expressed as statistical distributions representing prior beliefs about the original data generating processes, and (3) posterior distributions of these parameters, generated by computing the likelihood models and priors with the observed data [@wrightSystematistGuideEstimating2019]. The posterior represents the outcome of the model and includes distributions of plausible values for all of the parameters specified in the models, for example tree topologies and divergence times. Calculation of the posterior probabilities is typically too complex to integrate analytically using traditional methods, which is one reason why parsimony methods have been so dominant in phylogenetics. A key breakthrough that has recently made Bayesian phylogenetics more practical is efficient and user-friendly implementations of Markov Chain Monte Carlo (MCMC) sampling for computing the posterior probabilities. MCMC is a simulation technique that randomly draws on values from the priors to construct large numbers of different trees and models of evolution. Each link in the chain is a distinct description of the tree and a certain specification of all the parameters in the model of evolution [@holderPhylogenyEstimationTraditional2003]. Millions of trees are generated using different parameter values, and then scored according to their fit with the observed data, and when new parameter results in a better fit, it is used as the starting point for the next link in the chain [@wrightSystematistGuideEstimating2019]. At the end of the analysis, we get an estimate of the probability that any particular tree is the true evolutionary tree, given the observed data, and estimates of all the parameters of interest [@nascimentoBiologistGuideBayesian2017]. MCMC sampling for Bayesian evolutionary analysis has been implemented in free, open source, extensible and flexible software platforms such as BEAST2, MrBayes, and RevBayes, which are beginning to be explored by archaeologists [@gjesfjeld2019contributions; @buckley2017evolution]. In our case study below we introduce RevBayes because it is the only program with an interactive scripting language and has syntax very similar to the R language [@hohnaPhylogeneticInferenceUsing2017; @hohnaRevBayesBayesianPhylogenetic2016], which is advantageous because R is already widely used by archaeologists [@schmidtToolDrivenRevolutionsArchaeological2020].

```{r}
#| label: fig-tripartite-models
#| fig-cap: "Schematic of the key elements of the three models used in many palaeontological phylogenetic studies. Top row: Substitution models, with examples including binary and multistate models for discrete characters, and Brownian motion for continuous attributes. Middle row: Clock models, with examples including strict clock (with a constant rate among branches), uncorrelated relaxed clock (with a distinct rate along each branch drawn from a chosen probability distribution), and autocorrelated relaxed clock (with a distinct rate along each branch that is correlated with the rate along its parent branch). Bottom row: Tree models, include examples of pure birth (assumes speciation is constant over time, that we have no extinction), birth-death (where are species or variant disappear through extinction), and fossilized birth-death trees (incorporates the fossil recovery process and provides an expression for the probability of observing a tree with samples recovered along internal branches). Tree models allow specification of the speciation rate (λ), extinction rate (μ), and fossil recovery rate (ψ). Adapted from Warnock and Wright (2020), Symonds and Blomberg (2015), and Ho and Duchêne (2014)"

@@

In palaeontological phylogenetics there are three likelihood models that are the starting points for most analyses: the substitution model, the clock model, and the tree model [@warnockUnderstandingTripartiteApproach2020; @fig-tripartite-models]. Substitution models describe the process the researcher believes generated the character data. There are two main categories, each having several variants: discrete (e.g. Mk) and continuous (e.g. Brownian motion, Ornstein--Uhlenbeck, Lévy processes). Clock models describes the manner in which the researcher believes rates of evolution are distributed across the tree. There are two main groups of clock models: strict (with a constant rate among branches) and and relaxed, with examples including uncorrelated relaxed clock (with a distinct rate along each branch drawn from a chosen probability distribution), and autocorrelated relaxed clock (with a distinct rate along each branch that is correlated with the rate along its parent branch). Tree models describe the extinction, speciation, and sampling events that potentially led to the observed trees. Examples of tree models include pure birth (assumes speciation is constant over time, that we have no extinction), birth-death (where are species or variant disappear through extinction), and fossilized birth-death trees (incorporates the fossil recovery process and provides an expression for the probability of observing a tree with samples recovered along internal branches). The tripartite model is an important framework that enables nearly endless combinations of substitution, clock, and tree models to be assembled into a complete model.

@@

With the increased emphasis on modelling phylogenetic change in palaeontology came a concern that the discrete character substitution models for used to models change among nucleotides were poorly suited to modelling change in morphological characters. This concern motivated the development of methods to find more appropriate models and include morphological data without loss of information due to discretization. Simulations of palaeontological phylogenies using continuous traits have shown that models of stochastic Markovian movement through continuous trait space result in higher overall accuracy than discrete character models using discretized continuous characters [@parins-fukuchiUseContinuousTraits2018]. Traditional linear dimensions are widely used in palaeontological phylogenetics, but they are limited in the information they capture about shape and the interrelations among the dimensions of the studied structure. To overcome these limitations, palaeontologists have been using of geometric morphometrics to quantify shape as continuous characters to input into phylogenetic models. This includes landmark [@parins-fukuchiMorphologicalPhylogeographicEvidence2021; @alvarez-carreteroBayesianEstimationSpecies2019; @gonzalez-joseCladisticAnalysisContinuous2008] and outline [@smithGeometricMorphometricCharacter2013; @gabelaiaPhylogeneticSignalsScale2017; @lafumaMultipleEvolutionaryOrigins2021 ] methods to summarise shapes, and a variety of multivariate analyses to reduce dimensionality (e.g. PCA), accompanied by vibrant debate for [@gonzalez-joseMorphometricVariablesCan2011a; @zelditch2012geometric] and against [@varon-gonzalezEstimatingPhylogeniesShape2020; @klingenbergTestingQuantifyingPhylogenetic2010; @adamsMorphometricsPhylogeneticsPrincipal2011 ] their effectiveness.

```{r}
#| label: fig-efa-schematic
#| fig-cap: "..."

@@

Landmark and outline analysis of artefact shape is widely used in archaeology. Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone or metal tools [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019], or ceramic assemblages with distinct components [@Selden2019; @Topi2017]. For artefacts where landmarks are ambiguous or absent, outline approaches, such as Elliptic Fourier Analysis (EFA), are more effective for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Bonhomme2014]. Coordinates along a curve are converted into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. Outline analyses has long been effectively employed in the analysis of stone artefacts [e.g. @Gero1984; @Iovictua2009; @Fox2015; @Lipo2016; @Hoggard2019; @Ivanovaite2020; @riede2019reconciling; @matzig2021] and pottery [@Wilczek2014; @wang2020]. These studies share a standard approach of digitizing artefact shapes, extracting outlines, computing an EFA on these outlines, and then reducing the dimensionality of the harmonics by inputting them into a principal component analysis (PCA). The principal component (PC) scores are then interpreted as artefact shape variables that are typically more informative than linear dimensions. In our case study we extend the use of these artefact shape variables by using them as continuous characters to input into phylogenetic models.

@@

We demonstrate the use of artefact shape variables as continuous characters for phylogenetic modelling using flaked stone arrowheads found in late Neolithic chiefly burials across north-west France, southern Britain and Denmark during 2500 BC and 1700 BC [4500--3700 cal BP, @nicolasArrowsPowerBrittany2017]. These arrowheads, sometimes found in wooden boxes or wrapped in leather, are part of suite of grave goods that include bell-shaped ceramic food vessels, ornaments, and tools made from amber, bronze, copper, gold, silver and other materials. This set of grave goods is known as the Bell Beaker culture that spread across Europe through a mixture of migration and cultural diffusion. The arrowheads were bifacially flaked using antler, bone, and metal indentors, and many have traces of adhesive residue used for halfting onto a shaft. The usewear traces and geometry of halfting indicate that the arrowheads were loosely attached, designed for symbolic displace rather than efficient shots [@nicolasArrowsPowerBrittany2017]. Diagnostic impact fractures are generally absent from the arrowheads, with less than three percent showing signs of having been shot, indicating that most were not used as projectiles, but as non-functional display items. They were likely made by specialist master craft-workers, with the production sequence transmitted through generations, and the final products soley intended to portray and emphasise the social status of the individuals they were buried with. Given that the arrowheads were primarily symbols that were not used as projectiles, we can be confident that shape variation is unrelated to physical performance factors, and instead likely reflects transmission of social values and identities.

@@

We generated fours models to investigate the most appropriate combination of substitution, clock, and tree models. @tbl-model-summary summarises the model configurations and their marginal likelihoods. Full details of our Bayesian MCMC procedure are documented in our supplementary materials at http://doi.org/xxx/xxx. Taking the first model as the null model, the Bayes Factors indicate strong support for the model using a birth-death tree (where variants can appear and disappear, rather than only appear), relaxed Brownian motion (where the rate of morphological change is allowed to vary among branches of the phylogeny) and a relaxed clock (allowing for shifts in rates of morphological change across branches). @fig-treemap-densitree-map summarises the phylogenetic landscape, with the top left panel showing a single cluster of related trees, indicating that one phylogeny is definitively supported by the data [@jombartTreespaceStatisticalExploration2017]. The top right panel is a Densitree plot of trees randomly sampled from the converged posterior to provide intuition about the extent of incongruence among these trees [@bouckaertDensiTreeMakingSense2010]. It shows high congruence for the Bronze A2 and Bell Beaker samples, but substantial uncertainty for the Bronze A1 arrowheads, indicating competing topologies and possibly horizontal transmission and independent convergence on similar shapes.

@@

The middle and lower panel of @fig-treemap-densitree-map demonstrates the presence of a phylogenetic signal in our data. This is the tendency for related artefacts to resemble each other more than they resemble species drawn at random from the tree [@blombergTestingPhylogeneticSignal2003]. The middle panel shows the maximum a posteriori topology which is the tree that has the greatest posterior probability, averaged over all branch lengths and substitution parameter values. The middle panel also shows the strength of phylogenetic signal for each shape variable via the Local Indicator of Phylogenetic Association (local Moran's I) for each specimen and trait to indicate hotspots of association [@keckPhylosignalPackageMeasure2016]. The bottom panel of @fig-treemap-densitree-map show's Pagel's lambda values for our traits, nearly all are \~1 with p \< 0.05, indicating that there is a significant phylogenetic signal for all traits in our model, compared to a random model of evolution (lambda \~ 0).

```{r}
#| label: fig-treemap-densitree-map
#| fig-cap: "Exploration of landscapes of phylogenetic trees. Top left: Pairwise Euclidean distances between trees decomposed into a low-dimensional space using multidimensional scaling (MDS); top right: Densitree of all trees in the set drawn transparently, areas where a lot of the trees agree in topology and branch lengths show up as highly colored areas, while areas with little agreement show up as webs; middle panel:local Moran's I for each shape variable to detect hotspots of autocorrelation, red bars indicate significant I values (p < 0.01), indicating phylogenetic signal.; bottom panel: Pagel's lamba values for each shape variable, all have p < 0.05."
#| fig-height: 10
#| fig-width: 7
#| out-height: "500"

@@

The relaxed clock model provides insights into how rates of change in artefact shape varied across the tree. The left panel of @fig-tree-rates-morpho shows that rates of evolution are mostly fairly even across most the tree, with the slowest rates in an EBA-1 clade and the fastest rates in an EBA-2 clade. One interpretation of this is that the craft workers that made the earliest arrowheads were conservative in their exploration of shape possibilities as they were making their first appearance, then briefly experimented with a faster rate of change as use of the arrowheads became more established. Perhaps a conformist bias was at work as the arrowheads initially became established as burial goods. The right panel of @fig-tree-rates-morpho shows the distribution of trait values across the tree for the first five principal components, as our primary shape variables, and projections of the tree onto the morphospace for each combination of these variables. This shows that the deep tang (PC1) and narrow shoulders (PC2) were characteristics typical of the clades representing the middle of study period. We can also see that the area of the tree with highest rates of change has extreme values of (PC1: deep tang), indicating that much of the change occurred in that trait.

@@

Our brief case study demonstrates an accessible workflow using open source software tools for reconstructing phylogenetic relationships among archaeological artefacts. We have demonstrated how continuous characters, in the form of shape variables derived from geometric morphometry, which have become a popular way of quantitatively describing complex artefacts [@matzig2023], can be input into a phylogenetic model to yield easily interpretable output. The arrowheads of our case study are suitable because their shape did not change over the use-life of the artefact -- they were produced as grave goods and not used as projectiles. Other artefacts made using reductive techniques, such as Mousterian scrapers, might not be suitable for this kind of analysis because their shape changes substantially over their use-life [@dibbleInterpretationMiddlePaleolithic1987], independent of vertical transmission of cultural information about artefact shape.

Our use of Bayesian methods has provided formal approaches to manage competing models about the processes that generated the data, and methods for formalising uncertainty. These can be contrasted with previous work using parsimony methods that typically do not convey error estimates. However, Bayesian methods presents new challenges in the selection of appropriate models and prior parameter distributions. We anticipate that ethnographic and detailed archaeological data on transmission dynamics may help here.

This brief case study yielded promising results and sketches a straightforward template for future work in archaeological phylogenetics. We have demonstrated a model-based Bayesian workflow, and how to convey uncertainty, obtain insights about phylogenetic signal, and rates and locations of change. This case study could be extended with samples from a larger area, for example South Britain and Denmark @nicolasFlechesPouvoirAube2016], and by parallel analyses of other artefact classes found with the burials. More generally whole 'society' transmission dynamics or contrasting transmission across genders could also be explored with these methods. At much broader scales, phylogenetic/macroevolutionary studies of gene-culture co-evolution may be possible, as well as exploration of cultural phylogeography and niche construction dynamics using comparative methods.

@@

@@

@@

@@

Conflicts of Interest: Author A and Author B declare none

@@

>>>>

Bayesian inference consists of three parts: (1) likelihood models describing the processes that generated the data (statements of what the researcher believes to be true about their data), with the models containing (2) parameters expressed as statistical distributions representing prior beliefs about the original data generating processes, and (3) posterior distributions of these parameters, generated by computing the likelihood models and priors with the observed data [@wrightSystematistGuideEstimating2019]. The posterior represents the outcome of the model and includes distributions of plausible values for all of the parameters specified in the models, for example tree topologies and divergence times. Calculation of the posterior probabilities is typically too complex to integrate analytically using traditional methods, which is one reason why parsimony methods have been so dominant in phylogenetics. A key breakthrough that has recently made Bayesian phylogenetics more practical is efficient and user-friendly implementations of Markov Chain Monte Carlo (MCMC) sampling for computing the posterior probabilities. MCMC is a simulation technique that randomly draws on values from the priors to construct large numbers of different trees and models of evolution. Each link in the chain is a distinct description of the tree and a certain specification of all the parameters in the model of evolution [@holderPhylogenyEstimationTraditional2003]. Millions of trees are generated using different parameter values, and then scored according to their fit with the observed data, and when new parameter results in a better fit, it is used as the starting point for the next link in the chain [@wrightSystematistGuideEstimating2019]. At the end of the analysis, an estimate of the probability that any particular tree is the true evolutionary tree is generated, given the observed data, and estimates of all the parameters of interest [@nascimentoBiologistGuideBayesian2017]. MCMC sampling for Bayesian evolutionary analysis has been implemented in free, open source, extensible and flexible software platforms such as BEAST2, MrBayes, and RevBayes, which are beginning to be explored by archaeologists [@gjesfjeld2019contributions; @buckley2017evolution]. In our case study below we introduce RevBayes as it is currently the only program with an interactive scripting language and has syntax very similar to the R language [@hohnaPhylogeneticInferenceUsing2017; @hohnaRevBayesBayesianPhylogenetic2016], which is advantageous because R is already widely used by archaeologists [@schmidtToolDrivenRevolutionsArchaeological2020].

In palaeontological phylogenetics there are three likelihood models that are the starting points for most analyses: the substitution model, the clock model, and the tree model [@warnockUnderstandingTripartiteApproach2020; @fig-tripartite-models]. Substitution models describe the process the researcher believes generated the character data. There are two main categories, each having several variants: discrete (e.g. Mk) and continuous (e.g. Brownian motion, Ornstein--Uhlenbeck, Lévy processes). Clock models describes the manner in which the researcher believes rates of evolution are distributed across the tree. There are two main groups of clock models: strict (with a constant rate among branches) and relaxed, with examples including uncorrelated relaxed clock (with a distinct rate along each branch drawn from a chosen probability distribution), and autocorrelated relaxed clock (with a distinct rate along each branch that is correlated with the rate along its parent branch). Tree models describe the extinction, speciation, and sampling events that potentially led to the observed trees. Examples of tree models include pure birth (assumes speciation is constant over time, that we have no extinction), birth-death (where are species or variant disappear through extinction), and fossilized birth-death trees (incorporates the fossil recovery process and provides an expression for the probability of observing a tree with samples recovered along internal branches). The tripartite model is an important framework that enables nearly endless combinations of substitution, clock, and tree models to be assembled into a complete model.

```{r}
#| label: fig-tripartite-models
#| fig-cap: "Schematic of the key elements of the three models used in many palaeontological phylogenetic studies. Top row: Substitution models, with examples including binary and multistate models for discrete characters, and Brownian motion for continuous attributes. Middle row: Clock models, with examples including strict clock (with a constant rate among branches), uncorrelated relaxed clock (with a distinct rate along each branch drawn from a chosen probability distribution), and autocorrelated relaxed clock (with a distinct rate along each branch that is correlated with the rate along its parent branch). Bottom row: Tree models, include examples of pure birth (which assumes diversification is constant over time, that we have no extinction), birth-death (where taxa disappear through extinction), and fossilized birth-death trees (which incorporates the fossil recovery process and provides an expression for the probability of observing a tree with samples recovered along internal branches). Tree models allow specification of the diversification rate (λ), extinction rate (μ), and fossil recovery rate (ψ). Adapted from Warnock and Wright (2020), Symonds and Blomberg (2015), and Ho and Duchêne (2014)"

@@

@@

With the increased emphasis on modelling phylogenetic change in palaeontology came a concern that the discrete character substitution models for used to models change among nucleotides were poorly suited to modelling change in morphological characters. This concern motivated the development of methods to find more appropriate models and include morphological data without loss of information due to discretization. Simulations of palaeontological phylogenies using continuous traits have shown that models of stochastic Markovian movement through continuous trait space result in higher overall accuracy than discrete character models using discretized continuous characters [@parins-fukuchiUseContinuousTraits2018]. Traditional linear dimensions are widely used in palaeontological phylogenetics, but they are limited in the information they capture about shape and the interrelations among the dimensions of the studied structure -- as is also the case for lithic analysis [@Shott_Trail_2010]. To overcome these limitations, palaeontologists have been using of geometric morphometrics to quantify shape as continuous characters to input into phylogenetic models. This includes landmark [@parins-fukuchiMorphologicalPhylogeographicEvidence2021; @alvarez-carreteroBayesianEstimationSpecies2019; @gonzalez-joseCladisticAnalysisContinuous2008] and outline [@smithGeometricMorphometricCharacter2013; @gabelaiaPhylogeneticSignalsScale2017; @lafumaMultipleEvolutionaryOrigins2021 ] methods to summarise shapes, and a variety of multivariate analyses to reduce dimensionality (e.g. PCA), accompanied by vibrant debate for [@gonzalez-joseMorphometricVariablesCan2011a; @zelditch2012geometric] and against [@varon-gonzalezEstimatingPhylogeniesShape2020; @klingenbergTestingQuantifyingPhylogenetic2010; @adamsMorphometricsPhylogeneticsPrincipal2011 ] their effectiveness.

```{r}
#| label: fig-efa-schematic
#| fig-cap: "Schematic visualisation of the workflow proposed and used in this study. First, the outlines of 2D artefact images are described in so called harmonics using Elliptic Fourier Analysis, then these harmonics are then transformed and reduced in dimensions using a Principal Components Analysis. The Principal Components can be interpreted as artefact shape variables and be used as continuous character traits for the inference of phylogenies."

@@

Landmark and outline analysis of artefact shape is widely used in archaeology. Landmark-based morphometrics has been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone or metal tools [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019], or ceramic assemblages with distinct components [@Selden2019; @Topi2017]. For artefacts where landmarks are ambiguous or absent, outline approaches, such as Elliptic Fourier Analysis (EFA), are more effective for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Bonhomme2014]. Coordinates along a curve are converted into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. Outline analyses has long been effectively employed in the analysis of stone artefacts [e.g. @Gero1984; @Iovictua2009; @Fox2015; @Lipo2016; @Hoggard2019; @Ivanovaite2020; @riede2019reconciling; @matzig2021] and pottery [@Wilczek2014; @wang2020]. These studies share a standard approach of digitizing artefact shapes, extracting outlines, computing an EFA on these outlines, and then reducing the dimensionality of the harmonics by inputting them into a principal component analysis (PCA). The principal component (PC) scores are then interpreted as artefact shape variables that are typically more informative than linear dimensions. In our case study we extend the use of these artefact shape variables by using them as continuous characters to input into phylogenetic models [@fig-efa-schematic].

@@

We demonstrate the use of artefact shape variables as continuous characters for phylogenetic modelling using flaked stone arrowheads found in late Neolithic elite burials across north-west France, southern Britain and Denmark during 2500 BCE and 1700 BCE [4500--3700 cal BP, @nicolasArrowsPowerBrittany2017]. These arrowheads, sometimes found in wooden boxes or wrapped in leather, are part of a suite of grave goods that include bell-shaped ceramic food vessels, ornaments, and tools made from amber, bronze, copper, gold, silver and other materials. This rather distinct set of grave goods is characteristic of the Bell Beaker culture that spread across Europe through a mixture of migration and cultural diffusion. The arrowheads were bifacially flaked using antler, bone, and metal indentors, and many have traces of adhesive residue used for halfting onto a shaft. The usewear traces and geometry of halfting indicate that the arrowheads were loosely attached, designed for symbolic display rather than efficient shots [@nicolasArrowsPowerBrittany2017]. Diagnostic impact fractures are generally absent from the arrowheads, with less than three percent showing signs of having been shot, indicating that most were not used as projectiles, but as non-functional display items. They were likely made by specialist master craft-workers, with the production sequence transmitted through generations, and the final products solely intended to portray and emphasise the social status of the individuals they were buried with. Given that the arrowheads were primarily symbols that were not used as projectiles, we can be confident that shape variation is unrelated to physical performance factors or rejuvenation, and instead likely reflects transmission of social values and identities.

@@

In RevBayes, we generated fours models to investigate the most appropriate combination of substitution, clock, and tree models. @tbl-model-summary summarises the model configurations and their marginal likelihoods. Full details of our Bayesian MCMC procedure are documented in our supplementary materials at http://doi.org/xxx/xxx. Taking the first model as the null model, the Bayes Factors indicate strong support for the model using a birth-death tree (where variants can appear and disappear, rather than only appear), relaxed Brownian motion (where the rate of morphological change is allowed to vary among branches of the phylogeny) and a relaxed clock (allowing for shifts in rates of morphological change across branches). @fig-treemap-densitree-map summarises the phylogenetic landscape, with the top left panel showing a single cluster of related trees, indicating that one phylogeny is definitively supported by the data [@jombartTreespaceStatisticalExploration2017]. The top right panel is a Densitree plot of trees randomly sampled from the converged posterior to provide intuition about the extent of incongruence among these trees [@bouckaertDensiTreeMakingSense2010]. It shows high congruence for the Bronze A2 and Bell Beaker samples, but substantial uncertainty for the Bronze A1 arrowheads, indicating competing topologies and possibly horizontal transmission and independent convergence on similar shapes.

@@

The middle and lower panels of @fig-treemap-densitree-map demonstrates the presence of a phylogenetic signal in our data. This is the tendency for related artefacts to resemble each other more than they resemble specimens drawn at random from the tree [@blombergTestingPhylogeneticSignal2003]. The middle panel shows the maximum a posteriori topology which is the tree that has the greatest posterior probability, averaged over all branch lengths and substitution parameter values. The middle panel also shows the strength of phylogenetic signal for each shape variable via the Local Indicator of Phylogenetic Association (local Moran's I) for each specimen and trait to indicate hotspots of association [@keckPhylosignalPackageMeasure2016]. The bottom panel of @fig-treemap-densitree-map show's Pagel's lambda values for our traits, nearly all are \~1 with p \< 0.05, indicating that there is a significant phylogenetic signal for all traits in our model, compared to a random model of evolution (lambda \~ 0).

```{r}
#| label: fig-treemap-densitree-map
#| fig-cap: "Exploration of landscapes of phylogenetic trees. Top left: Pairwise Euclidean distances between trees decomposed into a low-dimensional space using multidimensional scaling (MDS); Top right: Densitree of all trees in the set drawn transparently, areas where a lot of the trees agree in topology and branch lengths show up as highly colored areas, while areas with little agreement show up as webs; Middle panel: local Moran's I for each shape variable to detect hotspots of autocorrelation, red bars indicate significant I values (p < 0.01), indicating phylogenetic signal; Bottom panel: Pagel's lamba values for each shape variable, all have p < 0.05."
#| fig-height: 10
#| fig-width: 7
#| out-height: "500"

@@

The relaxed clock model provides insights into how rates of change in artefact shape varied across the tree. The left panel of @fig-tree-rates-morpho shows that rates of evolution are mostly fairly even across most the tree, with the slowest rates in an EBA-1 clade and the fastest rates in an EBA-2 clade. One interpretation of this is that the craft workers that made the earliest arrowheads were conservative in their exploration of shape possibilities as they were making their first appearance, then briefly experimented with a faster rate of change as use of the arrowheads became more established. Perhaps a conformist bias was at work as the arrowheads initially became established as burial goods. The right panel of @fig-tree-rates-morpho shows the distribution of trait values across the tree for the first five principal components, as our primary shape variables, and projections of the tree onto the morphospace for each combination of these variables. This shows that the deep tang (PC1) and narrow shoulders (PC2) were characteristics typical of the clades representing the middle of study period. We can also see that the area of the tree with highest rates of change has extreme values of (PC1: deep tang), indicating that much of the change occurred in that trait. Inferences about the rate of change across such a phylogeny are particularly useful for inferring modes of transmission as speed of change is closely linked to teaching and learning pathways [@Tehrani_Riede_2008]

@@

Our  case study demonstrates an accessible workflow using open source software tools for reconstructing phylogenetic relationships among archaeological artefacts. We have demonstrated how continuous characters, in the form of shape variables derived from geometric morphometry, which have become a popular way of quantitatively describing complex artefacts [@matzig2023], can be input into a phylogenetic model to yield readily interpretable output. The arrowheads of our case study are suitable because their shape did not change over the use-life of the artefact -- they were produced as grave goods and not used as projectiles. Other artefacts made using reductive techniques, such as Mousterian scrapers, might not be suitable for this kind of analysis because their shape changes substantially over their use-life [@dibbleInterpretationMiddlePaleolithic1987], independent of vertical transmission of cultural information about artefact shape.

Our use of Bayesian methods has provided formal approaches to manage competing models about the processes that generated the data, and methods for formalising uncertainty. These can be contrasted with previous work using parsimony methods that typically do not convey error estimates. However, Bayesian methods present new challenges in the selection of appropriate models and prior parameter distributions. We anticipate that ethnographic and detailed archaeological data on transmission dynamics may help here.

This brief case study yielded promising results and offers a straightforward template for future work in archaeological phylogenetics. We have demonstrated a model-based Bayesian workflow, and how to convey uncertainty, obtain insights about phylogenetic signal, and rates and locations of change. This case study could be extended with samples from a larger area, for example South Britain and Denmark [@nicolasFlechesPouvoirAube2016], and by parallel analyses of other artefact classes found with the burials. More generally whole 'society' transmission dynamics or contrasting transmission across genders could also be explored with these methods using a wider array of different artefact classes. At much broader scales, phylogenetic/macroevolutionary studies of gene-culture co-evolution may be possible, as well as exploration of cultural phylogeography and niche construction dynamics using comparative methods.

@@

Earlier versions of this paper were presented at the Cultural Evolution Society Conference 2021, Sapporo, Japan, and the 13th International Symposium on Knappable Materials, 2021, Tarragona, Spain. We are grateful to the organisers of those events for the opportunity to share our work, and grateful to the participants who shared valuable feedback.

@@

BM, FR, DNM: Conceptualization, Methodology, Software. DNM: Data curation. BM: Writing- Original draft preparation, Visualization, Investigation. FR, DNM: Writing- Reviewing and Editing

@@

FR and DNM’s contribution is part of CLIOARCH, an ERC Consolidator Grant project which has received funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement No. 817564).

@@

The authors declare no conflicts of interest.

@@

To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in @Marwick2017, we include the entire R and RevBayes code used for all the analysis and visualizations contained in this paper in our online materials at http://doi.org/10.17605/OSF.IO/RBQA5. Also in this compendium are the raw data for all the tests reported here, and the outputs from the MCMC runs. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. In our online materials, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see @marwick2018packaging).
